Bayesian Sparsification of Deep $\cplx$-valued networks

### 2020-01-31 paper/paper.tex
placeholders for figures and some edits

In this section we study the complex variational dropout techniques discussed above
on the tasks and datasets, that were studied in prior research on complex-valued networks.
All experiments share the following settings unless specified otherwise:
\begin{itemize}
  \item ADAM optimizer with the base learning rate of $10^{-3}$
  \item learning rate the annealing schedule from \citep{trabelsi_deep_2017}
  \item global gradient clipping within an $\ell_2$-norm ball of radius $0.05$
\end{itemize}
Every experiment is replicated several times to take random effects into account, e.g.
random network initialization prior to ``dense'' stage, randomly ordered batches during
SGD. Throughout every experiment on MNIST-like datasets (sec.~\ref{sub:mnist_like_datasets},
we \textit{fixed} a random subset of the train split of each dataset, and used the test
split to gauge the final performance. Same approach was applied to CIFAR10 (sec.~%
\ref{sub:cifar10_100} experiments, which could partly explain the overall poor test
performance we got from the tested VGG16 model ($\real$ and $\cplx$ version).

@@

The goal of the following experiments to study the sparsity-performance trade-off of the
proposed Bayesian sparsification methods for $\cplx$-valued networks. To this end we employ
the following training plan in every experiment. The model is trained in three successive
stages: ``dense'' $\to$ ``sparsify'' $\to$ ``fine-tune''. Parameter values and the internal
state of the optimizer for similar parameters are kept between stages.

At the \textit{``dense''} stage every network is fit ``as-is'', retaining its original
non-bayesian architecture and using only the likelihood from \eqref{eq:elbo} without
variational dropout.
%
During the \textit{``sparsify''} stage we make every layer bayesian and apply $\cplx$-variational
dropout sec.~\ref{ssub:variational_dropout} (or sec.~\ref{ssub:empirical_bayes}) or its $\real$
counterpart. To be able to study the attainable sparsity levels we inject a coefficient $
  C \in (0, 1]
$ at the KL divergence term in \eqref{eq:elbo}:
\begin{equation}  \label{eq:elbo_with_coef}
    - \frac{C}N KL(q_{\theta} \| \pi_{\lambda})
    + \frac1{M} \sum_{i=1}^M
        \log p_{\phi}(z_i \mid g(\varepsilon_i; \theta))
    \,. \tag{\ref{eq:elbo}'}
\end{equation}
Higher $C$ yields increasingly more sparse layers upon termination of this stage.
%
At the \textit{``fine-tune''} stage we deduce sparsity masks from the parameters of the fit
variational approximation and train the remaining non-zero parameters similarly to the
``dense'' stage, i.e. with non-Bayesian layers, but with fixed sparsity masks.

Between \textit{``sparsify''} and \textit{``fine-tune''} stages we compute masks of non-zero
weights in each layer based on the relevance scores $\alpha$ (sec.~\ref{sub:dropout}). Since
the variational approximation factorizes into univariate distributions, whether a $\cplx$
or $\real$ parameter is zero or not is determined from individual relevance scores $
  \alpha = \tfrac{\sigma^2}{\lvert\mu\rvert^2}
$: non-zero if $\log \alpha \leq \tau$, zero otherwise. Figure~\ref{fig:log-relevance}
(top) displays general shape of the distribution of $\log \alpha$ values from the experiment
in sec.~\ref{sub:mnist_like_datasets}, which depicts the typical behaviour noticed in all
experiments conducted on every considered dataset: there always is a single mode within the
moderately negative range, and one or two modes to the right of $0$. The relative mass of
each mode changes with $C$ and the higher the coefficient is, the more mass is shifted to
the positive half. In every experiment the trough between the peaks of the negative mode
and the first positive one is around zero.

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.9\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[width=1.\linewidth,draft]{../notebooks/assets/log-alpha__histograms.pdf}
    % \caption{} \label{fig:log-relevance-kde}
  \end{subfigure} \\
  % \hfill
  \begin{subfigure}[b]{0.9\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[width=1.\linewidth,draft]{../notebooks/assets/proba_of_nonzero.pdf}
    % \caption{} \label{fig:log-relevance-proba}
  \end{subfigure}
  \caption{%
    \textit{top}: Estimate of $\log\alpha$ density for a convolutional and dense layer
    from the convolutional model of sec.~\ref{sub:mnist_like_datasets}.
    \textit{bottom}: The probability of a parameter being within $\pm \delta$ relative
    tolerance of its mean.
  }
  \label{fig:log-relevance}
\end{figure}

@@

Although image data is not $\cplx$ in nature \citet{trabelsi_deep_2017} study the applicability
of $\cplx$-valued networks to image classification (CIFAR-10, CIFAR-100, and a subset of
SVHN). Their approach is to either embed $\real$ in $\cplx$ (by $\Re z = 0$), or train
a smaller auxiliary network, that synthesizes the imaginary component from the raw channel
data. In our experiments we apply the two-dimensional Fourier Transform to get $\cplx$-valued
inputs. We concede that this is a rather unnatural representation of image data, however
the selection of publicly available datasets, which are not artificially $\cplx$-valued,
is limited to audio, speech and music.

Indeed, as is shown by the numerical experiments on MNIST-like datasets and CIFAR10, a
real-valued model trained on the Fourier transformed data underperforms relative to
its architectural twin, fit to on the raw colour data. At the same time complex-valued
models

% a couple of plots from MNIST and Fashion-MNIST
% write a note on the incompressible parameters (bias, batch norm etc.)
% discuss the plots describing how the test-performance grows with compression and then drops if the model is too sparse
% hinting at overfitting/regularization effects of masking and sparsification

@@

>>>>

In this section we study the complex variational dropout techniques discussed above
on the tasks and datasets, that were studied in prior research on complex-valued networks.
All experiments share the following settings unless specified otherwise:
\begin{itemize}
  \item ADAM optimizer with the base learning rate of $10^{-3}$
  \item learning rate the annealing schedule from \citep{trabelsi_deep_2017}
  \item global gradient clipping within an $\ell_2$-norm ball of radius $0.05$
\end{itemize}
Every experiment is replicated several times to take random effects into account, e.g.
random network initialization prior to training, randomly ordered batches during SGD.
Throughout every experiment on MNIST-like datasets (sec.~\ref{sub:mnist_like_datasets},
we \textit{fixed} a random subset of the train split of each dataset, and used the test
split to gauge the final performance. The same approach was applied to CIFAR10 (sec.~%
\ref{sub:cifar10_100} experiments, which could partly explain the overall poor test
performance we got from the tested VGG16 model ($\real$ and $\cplx$ version).

@@

The goal of the following experiments to study the sparsity-performance trade-off of the
proposed Bayesian sparsification methods for $\cplx$-valued networks. To this end we employ
the following training plan in every experiment. The model is trained in three successive
stages: ``dense'' $\to$ ``sparsify'' $\to$ ``fine-tune''. Parameter values and the internal
state of the optimizer for related parameters are transferred between stages.

At the \textit{``dense''} stage every network is fit ``as-is'', retaining its original
non-bayesian architecture and using only the likelihood term from \eqref{eq:elbo} without
variational dropout.
%
During the \textit{``sparsify''} stage we make every layer bayesian and apply $\cplx$-variational
dropout sec.~\ref{ssub:variational_dropout} (or sec.~\ref{ssub:empirical_bayes}) or its $\real$
counterpart. To be able to study the attainable sparsity levels we inject a coefficient $
  C \in (0, 1]
$ at the KL divergence term in \eqref{eq:elbo}:
\begin{equation}  \label{eq:elbo_with_coef}
    - \frac{C}N KL(q_{\theta} \| \pi_{\lambda})
    + \frac1{M} \sum_{i=1}^M
        \log p_{\phi}(z_i \mid g(\varepsilon_i; \theta))
    \,. \tag{\ref{eq:elbo}'}
\end{equation}
Higher $C$ yields increasingly more sparse layers upon termination of this stage.
%
The \textit{``fine-tune''} stage begins with computing the sparsity masks from the fitted
approximation $q_\theta(W)$ and proceeds to train the non-zero parameters similarly to
the ``dense'' stage, i.e. with non-Bayesian layers and fixed sparsity masks. The test
performance is reported for the model at the start and termination of the final stage.

Between \textit{``sparsify''} and \textit{``fine-tune''} stages we compute masks of non-zero
weights in each layer based on the relevance scores $\alpha$ (sec.~\ref{sub:dropout}). Since
the variational approximation factorizes into univariate distributions, whether a $\cplx$
or $\real$ parameter is zero or not is determined from individual relevance scores $
  \alpha = \tfrac{\sigma^2}{\lvert\mu\rvert^2}
$: non-zero if $\log \alpha \leq \tau$, zero otherwise. Figure~\ref{fig:log-relevance}
(top) displays the general shape of the distribution of $\log \alpha$ values from all
experiments in sec.~\ref{sub:mnist_like_datasets}. In fact, this appears to be the typical
in every experiment conducted in this study: there always is a single mode within the
moderately negative range, and one or two modes to the right of zero. The relative mass
of each mode changes with $C$ and the higher the coefficient is, the more mass is shifted
to the right. In every experiment the trough between the peaks of the negative mode and
the first positive one is around zero.

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.9\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[width=1.\linewidth,draft]{../assets/log-alpha__histograms.pdf}
    % \caption{} \label{fig:log-relevance-kde}
  \end{subfigure} \\
  % \hfill
  \begin{subfigure}[b]{0.9\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[width=1.\linewidth,draft]{../assets/proba_of_nonzero.pdf}
    % \caption{} \label{fig:log-relevance-proba}
  \end{subfigure}
  \caption{%
    \textit{top}: Estimate of $\log\alpha$ density for a convolutional and dense layer
    from the convolutional model of sec.~\ref{sub:mnist_like_datasets}.
    \textit{bottom}: The probability of a parameter being within $\pm \delta$ relative
    tolerance of its mean.
  }
  \label{fig:log-relevance}
\end{figure}

@@

Although image data is not $\cplx$ in nature \citet{trabelsi_deep_2017} study the applicability
of $\cplx$-valued networks to image classification (CIFAR-10, CIFAR-100, and a subset of
SVHN). Their approach is to either embed $\real$ in $\cplx$ (by $\Re z = 0$), or train
a smaller auxiliary network, that synthesizes the imaginary component from the raw channel
data. In our experiments we apply the two-dimensional Fourier Transform to get $\cplx$-valued
inputs. We concede that this is a rather unnatural representation of image data, however
the selection of publicly available datasets, which are not artificially $\cplx$-valued,
is limited to audio signals.

Indeed, as is shown by the numerical experiments on MNIST-like datasets and CIFAR10, a
real-valued model trained on the Fourier transformed data underperforms relative to
its architectural twin, fit to on the raw colour data. At the same time complex-valued
models.

% a couple of plots from MNIST and Fashion-MNIST
\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.9\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[width=1.\linewidth,draft]{../assets/log-alpha__histograms.pdf}
    % \caption{} \label{fig:log-relevance-kde}
  \end{subfigure} \\
  % \hfill
  \begin{subfigure}[b]{0.9\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[width=1.\linewidth,draft]{../assets/proba_of_nonzero.pdf}
    % \caption{} \label{fig:log-relevance-proba}
  \end{subfigure}
  \caption{%
    \textit{top}: The performance-compression tradeoff for MNIST \citep{citation_needed}.
    \textit{bottom}: The performance-compression tradeoff for Fashion MNIST \citep{citation_needed}.
  }
  \label{fig:mnist-like_tradeoff}
\end{figure}

% write a note on the incompressible parameters (bias, batch norm etc.)
% discuss the plots describing how the test-performance grows with compression and then drops if the model is too sparse
% hinting at overfitting/regularization effects of masking and sparsification

@@

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.9\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[width=1.\linewidth,draft]{../assets/log-alpha__histograms.pdf}
    % \caption{} \label{fig:log-relevance-kde}
  \end{subfigure} \\
  % \hfill
  \begin{subfigure}[b]{0.9\textwidth}  % imcl2019-style swears at this
    \centering
    \includegraphics[width=1.\linewidth,draft]{../assets/proba_of_nonzero.pdf}
    % \caption{} \label{fig:log-relevance-proba}
  \end{subfigure}
  \caption{%
    \textit{top}: MusicNet for the model of \citep{trabelsi_deep_2017} with kernel size $3$.
    \textit{bottom}: The performance-compression tradeoff for Fashion MNIST \citep{citation_needed}.
  }
  \label{fig:mnist-like_tradeoff}
\end{figure}
